Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-15 paper/sec-mva.tex
more edits

At the end of the cut-based analysis,
by combining the three event topologies,
we obtain a signal significance of $S/\sqrt{B}\simeq 1.7 (3.5)$
with all backgrounds (only QCD $4b$) taken into account.
%
We now show how this signal significance
 can be enhanced when the cut-based analysis
 is complemented by a multivariate analysis (MVA).
%
 Multivariate techniques are by now a mature tool
 in high-energy physics data analysis, opening
 new avenues for improving the performance
of many measurements and searches at high-energy colliders.
%
In particular, the classification of events into signal and
background processes by means of MVAs is a
commonly applied strategy in LHC
applications~\cite{Baldi:2014pta,Aaltonen:2012qt,
  Wardrope:2014kya,Chatrchyan:2013zna,Dall'Osso:2015aia}.

In this section, first of all we present the specific MVA that we use,
based on feed-forward multi-layer neural networks.
%
We then introduce the input variables that are
used in the MVA, including the jet substructure
variables, and then present the signal significance obtained
by applying the MVA.
%
Finally we assess the robustness of the MVA strategy in
the case of significant contamination from pileup.

@@

%
The specific type of  MVA that we use to
disentangle signal and background events is
a multi-layer feed-forward artificial neural network (ANN),
known as a {\it perceptron}.\footnote{This type of ANNs are the same
  as those used to parametrize Parton Distribution Functions
in the NNPDF global analyses~\cite{DelDebbio:2004qj,Ball:2008by,Ball:2011mu,Ball:2010de}.}
%
This family of ANNs are also known as {\it deep} neural networks, due to their multi-layered architecture.
%
The MVA inputs are a set of kinematic variables describing the
signal and background
events which satisfy the requirements of the
cut-based analysis.
%
The output of the trained ANNs also allows for the identification, in a fully automated way,
of the most relevant variables in the discrimination between 
signal and background.

@@

The training of the ANN for the signal/background classification task
proceeds as follows.
%
Given a set of $N_{\mathrm{var}}$  kinematic variables $\{k\}_i$ associated with the event $i$, and a set of neural network weight
parameters $\{\omega\}$, we interpret the neural network output $y_i$
(the activation state of the
neuron in the last layer)
as the probability that the event $i$ originates from the signal process,
\be
y_i = P(y^\prime_i=1|\{k\}_i, \{\omega\} )\, ,
\ee
where $y_i^\prime$ represents the true classification of the event $i$, {\it i.e},
$y^\prime = 1$ for signal and $y^\prime = 0$ for background events.
%
With this interpretation, our general classification probability including background events is given by
\be
P(y_i^\prime|\{k\}_i, \{\omega\}) = y_i^{y^\prime_i}(1-y_i)^{1-y^\prime_i} \, ,
\ee
consequently we can define an error function $E(\{\omega\})$
to be minimized during the ANN training. In this case, the error function is
the cross-entropy function, defined as
 \bea
 &&E(\{\omega\}) \equiv -\log\left(\prod_i^{N_{\text{ev}}} P(y_i^\prime|\{k\}_i, \{\omega\})\right)\nonumber\\
 &&=
 \sum_i^{N_{\text{ev}}} \lc y^\prime_i\log{y_i} + (1-y^\prime_i)\log{(1-y_i)}\rc \, ,
 \label{cross-entropy}
 \eea
 where $N_{\text{ev}}$ is the number of
 Monte Carlo events that are used for the ANN training.
 %
 The ANN is trained both on the signal and background MC events,
 so it is crucial to ensure that the input MC sample is large enough
 to avoid contamination from MC statistical fluctuations.

@@

 In order to avoid over-fitting,
 we have used a cross-validation stopping
 criterion, in particular the same one as
 that used in the NNPDF3.0 analysis~\cite{Ball:2014uwa}.
 %
 This cross-validation proceeds by dividing the input MC dataset into two disjoint sets,
 using one for training the ANN and the other for validation: the optimal
 stopping point is then given by the minimum of the error function
 Eq.~(\ref{cross-entropy}) to the validation sub-sample.
 %
 This indicates the point where we are training on the statistical fluctuations
 of the input MC samples, rather than in the underlying (smooth) physical
 distributions.

@@

 To ensure that we include the variables with the most discriminating power,
 we need to train
 the MVA on a broad array of kinematic variables.
%
In this work we use different sets of
input variables for the three categories.
%
In particular, in the case of large-$R$ jets (for the boosted
and intermediate categories), we are fully exploiting the available information 
on jet substructure.

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$.
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in the $\phi$--$\eta$ plane
  between the two Higgs candidates, $\Delta R_{hh}$.
  \item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\end{itemize}
In addition, in the boosted category we use
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,2}}$, Higgs candidate AKT03 subjets.
  %
  In the resolved category instead,
  the corresponding variables are
  the transverse momenta $p_{T,i}$ of the four leading 
  $b$-tagged small-$R$ jets in the event.
  %
  In the intermediate category, we use the
  transverse momenta of the subjets
  from the large-$R$ jet $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and the
 transverse momenta $p_{T,i}$ of the two leading 
  $b$-tagged small-$R$ jets.
  %
 Therefore, we have 13 variables which are common to the three categories.

>>>>

At the end of the cut-based analysis,
by combining the three event topologies,
we obtain a signal significance of $S/\sqrt{B}\simeq 1.8~(3.5)$
with all backgrounds (only QCD $4b$) taken into account.
%
We now show how this signal significance
 can be enhanced when the cut-based analysis
 is complemented by a multivariate analysis.
%
 Multivariate techniques are by now a mature tool
 in high-energy physics data analysis, opening
 new avenues for improving the performance
of many measurements and searches at high-energy colliders.
%
In particular, the classification of events into
signal and
background processes by means of MVAs is
commonly used in LHC
applications~\cite{Baldi:2014pta,Aaltonen:2012qt,
  Wardrope:2014kya,Chatrchyan:2013zna,Dall'Osso:2015aia}.

In this section, first of all we present the specific MVA that we use,
based on feed-forward multi-layer neural networks.
%
We then introduce the input variables that are
used in the MVA, including the jet substructure
variables, and then present the signal significance obtained
by applying the MVA.
%
Then we assess the robustness of the MVA strategy in
the case of significant contamination from pileup.

@@

%
The specific type of  MVA that we use to
disentangle signal and background events is
a multi-layer feed-forward artificial neural network (ANN),
known as a {\it perceptron}.\footnote{This type of ANNs are the same
  as those used to parametrize Parton Distribution Functions
in the NNPDF global analyses~\cite{DelDebbio:2004qj,Ball:2008by,Ball:2011mu,Ball:2010de}.}
%
This family of ANNs are also known as {\it deep neural networks},
due to their multi-layered architecture.
%
The MVA inputs are a set of kinematic variables describing the
signal and background
events which satisfy the requirements of the
cut-based analysis.
%
The output of the trained ANNs also allows for the identification, in a fully automated way,
of the most relevant variables in the discrimination between 
signal and background.

@@

The training of the ANN for the signal/background classification task
proceeds as follows.
%
Given a set of $N_{\mathrm{var}}$  kinematic variables $\{k\}_i$ associated with the event $i$, and a set of neural network weight
parameters $\{\omega\}$, we interpret the neural network output $y_i$
(the activation state of the
neuron in the last layer)
as the probability that the event $i$ originates from the signal process,
\be
y_i = P(y^\prime_i=1|\{k\}_i, \{\omega\} )\, ,
\ee
where $y_i^\prime$ represents the true classification of the event $i$, {\it i.e},
$y^\prime_i = 1$ for signal and $y^\prime_i = 0$ for background events.
%
With this interpretation, our general classification probability including background events is given by
\be
P(y_i^\prime|\{k\}_i, \{\omega\}) = y_i^{y^\prime_i}(1-y_i)^{1-y^\prime_i} \, ,
\ee
consequently we can define an error function $E(\{\omega\})$
to be minimized during the ANN training. In this case, the error function is
the cross-entropy function, defined as
 \bea
 &&E(\{\omega\}) \equiv -\log\left(\prod_i^{N_{\text{ev}}} P(y_i^\prime|\{k\}_i, \{\omega\})\right)\nonumber\\
 &&=
 \sum_i^{N_{\text{ev}}} \lc y^\prime_i\log{y_i} + (1-y^\prime_i)\log{(1-y_i)}\rc \, ,
 \label{cross-entropy}
 \eea
 where $N_{\text{ev}}$ is the number of
 Monte Carlo events that are used for the ANN training.
 %
 The ANN is trained both on the signal and background MC events,
 so it is important to ensure that the input MC sample is large enough
 to avoid contamination from MC statistical fluctuations.

@@

 In addition,
 in order to avoid the possibility of over-fitting,
 we have used a cross-validation stopping
 criterion, in particular the same one as
 that used in the NNPDF3.0 analysis~\cite{Ball:2014uwa}.
 %
 This cross-validation proceeds by dividing the input MC dataset into two disjoint sets,
 using one for training the ANN and the other for validation: the optimal
 stopping point is then given by the minimum of the error function
 Eq.~(\ref{cross-entropy}) to the validation sub-sample.
 %
 This indicates the point where
the ANN begins to train upon  statistical fluctuations
in the input MC samples, rather than learning
the underlying (smooth) physical  distributions.

@@

In this work we use different sets of
input variables for the three categories.
%
In the case of large-$R$ jets, we  exploit the available
information  on jet substructure.
%
For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$.
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in the $\phi$--$\eta$ plane
  between the two Higgs candidates, $\Delta R_{hh}$.
  \item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\end{itemize}
In addition, in the boosted category we use
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,2}}$, Higgs candidate AKT03 subjets.
  %
  In the resolved category instead,
  the corresponding variables are
  the transverse momenta $p_{T,i}$ of the four leading 
  $b$-tagged small-$R$ jets in the event.
  %
  In the intermediate category, we use the
  transverse momenta of the subjets
  from the large-$R$ jet $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and the
 transverse momenta $p_{T,i}$ of the two leading 
  $b$-tagged small-$R$ jets.
  %
 Therefore, we have 13 variables which are common to the three categories.
